Exponential Family Markets for Information Aggregation

### 2014-02-12 WriteUp/maxentpredmkts.tex
addressed some comments

%\subsection{Exponential Smoothing}
%One way to model the exponential family prediction market might be through exponential smoothing.
%\begin{itemize}
%\item A \emph{time series} is a set of data points sampled at periodic instances.
%\item \emph{Smoothing} is the process of creating an approximation function based on time series that while capturing  patterns in the data is relatively insensitive to noise.
%\item Natural first step is to use moving average. But average on how many samples (say $k$)? What to do until the first $k$ samples have been received?
%\item
%Next attempt might be to use weighted moving average. Advantage is that you can give higher weight to more recent terms. But still same disadvantage as simple moving average technique.
%\item
%\emph{Exponential Smoothing} attempts to remove this disadvantage.
%\end{itemize}
%
%Exponential smoothing is defined iteratively for $\alpha\in(0,1)$ as
%\begin{eqnarray*}
%s_{1} &=& x_{0}\\
%s_{t}&=&\alpha x_{t-1}+(1-\alpha)s_{t-1}
%\end{eqnarray*}
%
%Smaller values of $\alpha$ means greater smoothing e.g., $\alpha=0$ gives a constant function ($=x_{0}$).
%Note that the selection of $x_{0}$ is important -- and increases in importance as $\alpha$. Some advantages over moving average: all data points matter (decreasing importance with time) and computationally only last data point need be stored.
%\pagebreak
\section{Maximum Entropy Prediction Markets via Exponential Families}
\label{sec:maxentmarkets}

@@

A risk-neutral agent will choose to acquire shares up to the point where, for each share, expected payoff equals marginal price. Formally, if the agent acquires portfolio $\delta$, moving the market state vector to $\theta' = \theta + \delta$, then we must have
%
\begin{equation} \label{risk-neutral-agent}
\Exp_p[\phi(x)] = \grad C(\theta').
\end{equation}
%
\sindhu{sebastien/jake, should this be $\Exp_p(x;\theta')[\phi(x)]$?}
In this way, by its choice of $\delta$, the agent reveals that its expected belief is $\mu = \grad C(\theta')$. We stress that this observation relies on the assumptions that 1) the agent is risk-neutral, 2) the agent does not incorporate the market's information into its own beliefs, and 3) the agent is not budget constrained. We will examine relaxations of each assumption in later sections.

@@

The exponential families framework is well-suited to reasoning about Bayesian updates. As before let the data distribution be given by $p(x;\theta)=\exp(\la\theta,\phi(x)\ra-T(\theta))$ where $T$ is the log partition function and $\phi$ are the sufficient statistics. Instead of direct beliefs about the data distribution the agent maintains a conjugate prior over the parameters $\theta$. Every exponential family admits a conjugate prior of the form
%
$$p(\theta;b_{0})=\exp(\la n\nu,\theta\ra + nT(\theta) - \psi(\nu,n)).$$
%
Note that this is also an exponential family with natural parameter $b_0 = (n\nu,n)$ where $\nu \in \bR^d$ and $n$ is a positive integer. The sufficient statistic maps $\theta$ to $(\theta, T(\theta))$, and the log partition function $\psi$ is defined as the normalizer as usual. For a complete treatment of exponential families conjugate priors, see for instance~\citet{Barndorff78}. Now~\citet[Thm.\ 2]{diaconis79} and~\citet{jewell1974credible} have shown that
%
\begin{equation}\label{eq:prior}
\E_{\theta\sim b_{0}}\E_{x\sim \theta}[\phi(x)]=\nu,
\end{equation}
%
\sindhu{Maybe this is nitpicking but $\E_{\theta\sim p(\theta;b_{0})}\E_{x\sim p(x;\theta)}[\phi(x)]$?}
meaning that $\nu = n \nu / n$ is the posterior mean. Thus, it is helpful to think of the prior as being based on a `phantom' sample of size $n$ and mean $\nu$. Suppose now that the agent observes an empirical sample with mean $\hmu$ and size $m$. By a standard derivation~\citep[see][]{diaconis79}, the posterior conjugate prior parameters become $n\nu \leftarrow n\nu + m\hmu$ and $n \leftarrow n+m$, and the posterior expectation~(\ref{eq:prior}) evaluates to
%
\begin{equation} \label{posterior}
\frac{n\nu + m\hmu}{n+m}.
\end{equation}
%
Thus the posterior mean is a convex combination of the prior and empirical means, and their relative weights depend on the phantom and empirical sample sizes.% need a reference here

@@

\section{Risk-Averse Traders with Exponential Utility} \label{sec:exputil}
\sindhu{Sebastien, do you think we could cite \cite{norstad99util} for instance?}

In this section we relax the standard assumption that agents in the market are risk-neutral. We show that, with sufficient extra structure to the agents' beliefs and utilities, the market performs a clean aggregation of the agents' expectations in the form of a simple weighted average.
Assume that the agent has an exponential utility function for wealth $w$:
%
\begin{equation} \label{exp-util}
U_a(w) = -\frac{1}{a} \exp(-aw).
\end{equation}
%
Here $a$ controls the level of risk aversion: the agent is more risk averse as $a$ increases, and as $a$ tends to 0 we approach linear utility (risk-neutrality). Specifically, $a$ is the Arrow-Pratt coefficient of absolute risk aversion, and exponential utilities of the form~(\ref{exp-util}) are the unique utilities that exhibit constant absolute risk aversion (CARA).

@@

In the context of the generalized LMSR we write $T$ rather than $C$ where $T$ is the log partition function, with liquidity-adjusted version $T_{\lambda}$. Let $\hmu$ be the agent's mean belief with corresponding natural parameter $\htheta = \grad T^{-1}(\hmu)$. Recall that a risk-neutral agent moves the share vector so that the prices match its mean parameter. Therefore, define the \emph{target shares} as $\ttheta = \grad T^{-1}_{\lambda}(\hmu)$. The target shares $\htheta$ and natural parameter $\ttheta$ are related by
%
$$
\grad T(\htheta) = \grad T_{\lambda}(\ttheta) = \hmu.
$$
%
It is straightforward to check that $\grad T_{\lambda}(\ttheta) = \grad T(\lambda \ttheta)$ so we have
%
\begin{equation} \label{target}
\ttheta = \htheta / \lambda.
\end{equation}
%
Higher price responsiveness means fewer shares must be bought to make the market prices match the agent's expectation, so the natural parameter is scaled down accordingly. With a liquidity adjustment, the analysis of Theorem~\ref{thm:exp-util} can be extended and yields the following result.
%
\begin{corollary}\jake{$\mu$ without a hat is never defined... (this is for sebastien)}
Under the conditions of Theorem~\ref{thm:exp-util} and an inverse liquidity of $\lambda$, the agent's optimal trade $\delta$ moves the state vector to
%
\begin{equation} \label{liqu-update}
\theta + \delta \:\:=\:\: \frac{\lambda}{\lambda+a} \ttheta + \frac{a}{\lambda+a} \theta
\:\:=\:\: \frac{\lambda}{\lambda+a} \grad T^{-1}_{\lambda}(\hmu) + \frac{a}{\lambda+a} \grad T^{-1}_{\lambda}(\mu).
\end{equation}
%
\end{corollary}
%
According to~(\ref{liqu-update}), as $\lambda$ grows large the agent moves the market state closer to the \emph{target shares}, rather than its true natural parameter. Note that the target shares themselves depend on $\lambda$ by~(\ref{target}), but the update can be directly written in terms of the agent's beliefs as in the right-hand side of~(\ref{liqu-update}).

>>>>

%\subsection{Exponential Smoothing}
%One way to model the exponential family prediction market might be through exponential smoothing.
%\begin{itemize}
%\item A \emph{time series} is a set of data points sampled at periodic instances.
%\item \emph{Smoothing} is the process of creating an approximation function based on time series that while capturing  patterns in the data is relatively insensitive to noise.
%\item Natural first step is to use moving average. But average on how many samples (say $k$)? What to do until the first $k$ samples have been received?
%\item
%Next attempt might be to use weighted moving average. Advantage is that you can give higher weight to more recent terms. But still same disadvantage as simple moving average technique.
%\item
%\emph{Exponential Smoothing} attempts to remove this disadvantage.
%\end{itemize}
%
%Exponential smoothing is defined iteratively for $\alpha\in(0,1)$ as
%\begin{eqnarray*}
%s_{1} &=& x_{0}\\
%s_{t}&=&\alpha x_{t-1}+(1-\alpha)s_{t-1}
%\end{eqnarray*}
%
%Smaller values of $\alpha$ means greater smoothing e.g., $\alpha=0$ gives a constant function ($=x_{0}$).
%Note that the selection of $x_{0}$ is important -- and increases in importance as $\alpha$. Some advantages over moving average: all data points matter (decreasing importance with time) and computationally only last data point need be stored.
%\pagebreak
\section{Exponential Family Markets}
\label{sec:maxentmarkets}

@@

A risk-neutral agent will choose to acquire shares up to the point where, for each share, expected payoff equals marginal price. Formally, if the agent acquires portfolio $\delta$, moving the market state vector to $\theta' = \theta + \delta$, then we must have
%
\begin{equation} \label{risk-neutral-agent}
\Exp_p[\phi(x)] = \grad C(\theta').
\end{equation}
%
In this way, by its choice of $\delta$, the agent reveals that its expected belief is $\mu = \grad C(\theta')$. We stress that this observation relies on the assumptions that 1) the agent is risk-neutral, 2) the agent does not incorporate the market's information into its own beliefs, and 3) the agent is not budget constrained. We will examine relaxations of each assumption in later sections.

@@

The exponential families framework is well-suited to reasoning about Bayesian updates. As before let the data distribution be given by $p(x;\theta)=\exp(\la\theta,\phi(x)\ra-T(\theta))$ where $T$ is the log partition function and $\phi$ are the sufficient statistics. Instead of direct beliefs about the data distribution the agent maintains a conjugate prior over the parameters $\theta$. Every exponential family admits a conjugate prior of the form
%
$$p(\theta;b_{0})=\exp(\la n\nu,\theta\ra + nT(\theta) - \psi(\nu,n)).$$
%
Note that this is also an exponential family with natural parameter $b_0 = (n\nu,n)$ where $\nu \in \bR^d$ and $n$ is a positive integer. The sufficient statistic maps $\theta$ to $(\theta, T(\theta))$, and the log partition function $\psi$ is defined as the normalizer as usual. For a complete treatment of exponential families conjugate priors, see for instance~\citet{Barndorff78}. Now~\citet[Thm.\ 2]{diaconis79} and~\citet{jewell1974credible} have shown that
%
\begin{equation}\label{eq:prior}
\E_{\theta\sim b_{0}}\E_{x\sim \theta}[\phi(x)]=\nu,
\end{equation}
%
meaning that $\nu = n \nu / n$ is the posterior mean. Thus, it is helpful to think of the prior as being based on a `phantom' sample of size $n$ and mean $\nu$. Suppose now that the agent observes an empirical sample with mean $\hmu$ and size $m$. By a standard derivation~\citep[see][]{diaconis79}, the posterior conjugate prior parameters become $n\nu \leftarrow n\nu + m\hmu$ and $n \leftarrow n+m$, and the posterior expectation~(\ref{eq:prior}) evaluates to
%
\begin{equation} \label{posterior}
\frac{n\nu + m\hmu}{n+m}.
\end{equation}
%
Thus the posterior mean is a convex combination of the prior and empirical means, and their relative weights depend on the phantom and empirical sample sizes.% need a reference here

@@

\section{Risk-Averse Traders with Exponential Utility} \label{sec:exputil}

In this section we relax the standard assumption that agents in the market are risk-neutral. We show that, with sufficient extra structure to the agents' beliefs and utilities, the market performs a clean aggregation of the agents' expectations in the form of a simple weighted average.
Assume that the agent has an exponential utility function for wealth $w$:
%
\begin{equation} \label{exp-util}
U_a(w) = -\frac{1}{a} \exp(-aw).
\end{equation}
%
Here $a$ controls the level of risk aversion: the agent is more risk averse as $a$ increases, and as $a$ tends to 0 we approach linear utility (risk-neutrality). Specifically, $a$ is the Arrow-Pratt coefficient of absolute risk aversion, and exponential utilities of the form~(\ref{exp-util}) are the unique utilities that exhibit constant absolute risk aversion~\citep[Chap.\ 11]{VarianMicro}.

@@

In the context of the generalized LMSR we write $T$ rather than $C$ where $T$ is the log partition function, with liquidity-adjusted version $T_{\lambda}$. Let $\hmu$ be the agent's mean belief with corresponding natural parameter $\htheta = \grad T^{-1}(\hmu)$. Recall that a risk-neutral agent moves the share vector so that the prices match its mean parameter. Therefore, define the \emph{target shares} as $\ttheta = \grad T^{-1}_{\lambda}(\hmu)$. The target shares $\htheta$ and natural parameter $\ttheta$ are related by
%
$$
\grad T(\htheta) = \grad T_{\lambda}(\ttheta) = \hmu.
$$
%
It is straightforward to check that $\grad T_{\lambda}(\ttheta) = \grad T(\lambda \ttheta)$ so we have
%
\begin{equation} \label{target}
\ttheta = \htheta / \lambda.
\end{equation}
%
Higher price responsiveness means fewer shares must be bought to make the market prices match the agent's expectation, so the natural parameter is scaled down accordingly. With a liquidity adjustment the analysis of Theorem~\ref{thm:exp-util} can be extended and yields the following result, where as before $\theta$ and $\mu$ are the market's outstanding shares and prices respectively.
%
\begin{corollary}
Under the conditions of Theorem~\ref{thm:exp-util} and an inverse liquidity of $\lambda$, the agent's optimal trade $\delta$ moves the state vector to
%
\begin{equation} \label{liqu-update}
\theta + \delta \:\:=\:\: \frac{\lambda}{\lambda+a} \ttheta + \frac{a}{\lambda+a} \theta
\:\:=\:\: \frac{\lambda}{\lambda+a} \grad T^{-1}_{\lambda}(\hmu) + \frac{a}{\lambda+a} \grad T^{-1}_{\lambda}(\mu).
\end{equation}
%
\end{corollary}
%
According to~(\ref{liqu-update}), as $\lambda$ grows large the agent moves the market state closer to the \emph{target shares}, rather than its true natural parameter. Note that the target shares themselves depend on $\lambda$ by~(\ref{target}), but the update can be directly written in terms of the agent's beliefs as in the right-hand side of~(\ref{liqu-update}).
